---
title: "gpirt Vignette"
author: JBrandon Duck-Mayr and T. Ryan Johnson
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gpirt Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
bibliography: vignette-references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `gpirt` package provides an MCMC sampler and related tools for a
**G**aussian **P**rocess **I**tem **R**esponse **T**heoretic (GP-IRT) model.
Unlike other IRT models that make strong assumptions about the functional form
of the "item response function," or the function mapping respondents' latent
trait to their response to items, we place a GP prior over the distribution of
functions mapping latent traits to responses, and simulataneously learn these
functions and the respondents' latent traits.

For readers unfamiliar with Gaussian processes, we refer them @RW:2006,
available [here](http://www.gaussianprocess.org/gpml/chapters/).
In what follows, we first detail our MCMC algorithm for sampling the model
parameters' posterior distribution.
We then walk through a simple but complete example, from data preparation, to
prior specification and sampling, and analysis of results.

# Algorithm

We include at the outset, a brief explanation of the notation used throughout:

- $n$ is the number of respondents, indexed by $i$.
- $m$ is the number of items, indexed by $j$.
- $T$ is the number of iterations, indexed by $t$.
- $y$ is an $n \times m$ dimension matrix of each respondent's response to each
  item.
- $\theta$ is the $n \times d$ dimension matrix of respondents' latent traits,
  which we will commonly call ideology (as would be the case in the authors'
  substantive applications -- political science). Currently the algorithm is
  only implemented for $d = 1$.
- $f_j$ is the latent function mapping ideology to responses for item $j$,
  through a sigmoid function; i.e., $\Pr(y_{ij} = 1) = \sigma(f_j(\theta_i))$.
- $f = \lbrace f_{j} \rbrace$.
  In our code, we refer to the $n \times m$ matrix storing the current draw of
  $\lbrace f_j(\theta) \rbrace$ as `f`.

Superscripts denote iterations of the algorithm and subscripts denote
respondents/items.
For example, $\theta_i^t$ is the value drawn for respondent $i$'s ideology
parameter at iteration $t$ of our algorithm.

## Priors

We place a normal prior over each $\theta_i$;
the variance for all ideology parameter priors is 1,
but the prior means may vary.
Specifically, we assume (at least some of) the respondents can be classified
into different groups (in our political science application, parties),
and allow the user to specify a prior mean $\mu_g$ for each group $g$.
If a group or respondent is not assigned a prior mean by the user,
we assume $\mu_g = 0$ for that $g$.

We place a GP prior over the $f_j$, with a mean function of zero,
and covariance given by the squared exponential covariance function
(also called the radial basis function)

$$
K(\theta, \theta') = \sigma^2
\exp\left(-\frac{\| \theta - \theta' \|^2}{2\ell^2}\right),
$$

where $\ell$ is the *length scale*,
governing how quickly $f_j$ can change,
and $\sigma$ is the *scale factor*, or the function variance of $f_j$.

## MCMC Algorithm

Our MCMC algorithm proceeds as follows:

1. Initialize $f$ and $\theta$: $(f^{0}, \theta^{0})$.
    a. We allow users to specify initial values for $\theta$; otherwise,
       $\theta^0$ is drawn from the prior distributions of the $\theta_i$.
    b. Draw the $f_j$ from their prior distribution, i.e. a normal distribution
       with mean zero and covariance $K(\theta^0, \theta^0)$.
2. For $t = 1, ..., T$:
    a. Use an elliptical slice sampler to draw $f^t$ given $\theta^{t-1}$,
       $f^{t-1}$, and $y$. For more details on elliptical slice samplers,
       please consult @ESS,
       available [here](https://homepages.inf.ed.ac.uk/imurray2/pub/10ess/).
    b. Given $f^t$ and $\theta^{t-1}$, draw $f^* = \{ f_j(\theta^*) \}$
       for $\theta^* = \{-5, -4.99, \ldots, 4.99, 5\}$ from
       $$
       f_j^t(\theta^*) \sim \mathcal{N}(
         K(\theta^*, \theta^{t-1})\, K(\theta^{t-1}, \theta^{t-1})^{-1}f_j^t,\
         K(\theta^*, \theta^*) - 
         K(\theta^*, \theta^{t-1})\,
         K(\theta^{t-1}, \theta^{t-1})^{-1}\,
         K(\theta^{t-1}, \theta^*)
         )
       $$
       (see @RW:2006, equation 2.19).
    c. Use inverse transform sampling to draw $\theta^t$ by brute-force
       computation of the conditional posterior on the grid $\theta^*$.

# Example

We will use an example from the U.S. Senate.
Congressional voting data is available from [Voteview](https://voteview.com/)
[@voteview-data].
First let's download the data:

```{r get_data}
base_url  <- "https://voteview.com/static/data/out/"
vote_data <- read.csv(paste0(base_url, "votes/S116_votes.csv"),
                      stringsAsFactors = FALSE)
```

The data from Voteview is in a tidy format,^[See `tidyr`'s "Tidy Data" Vignette]
but because we want a matrix of responses, we will actually have to make it messy.
I'll present both a base R method of doing so, as well as a commented-out
`tidyverse` solution:

```{r reshape_data}
item_ids   <- unique(vote_data$rollnumber)
member_ids <- unique(vote_data$icpsr)
n          <- length(member_ids)
m          <- length(item_ids)
responses  <- matrix(nrow = n, ncol = m, dimnames = list(member_ids, item_ids))
for ( j in 1:m ) {
    item_idx        <- which(vote_data$rollnumber == item_ids[j])
    responses[ , j] <- vote_data$cast_code[item_idx]
}
## Tidyverse solution:
# library(dplyr)
# library(tidyr)
# responses <- vote_data %>%
#     ## We only need the vote identifier, member identifier, and vote code:
#     select(icpsr, rollnumber, cast_code) %>%
#     ## Then we spread the data out:
#     spread(icpsr, cast_code)
# rownames(responses) <- responses$rollnumber
# responses <- t(responses[ , -1])
```

Now, recall that for our $\theta_i$ priors, we can specify prior means by group.
This is useful for identifying the directionality of the space, particularly
when as in politics, we can readily say which groups should be on the "left"
and which groups should be on the "right" (though note that with a standard
deviation of one, this certainly does not outright prevent those in the "left"
group from ending up with a positive ideology estimate, and vice versa).
So, we use information about the Senators' party affilliation to specify
their group and create a `response_matrix` object:

```{r make_response_matrix}
member_data <- read.csv(paste0(base_url, "members/S116_members.csv"),
                        stringsAsFactors = FALSE)
icpsrs      <- as.character(member_data$icpsr)
party_codes <- member_data$party_code[match(member_ids, icpsrs)]
library(gpirt)
responses   <- as.response_matrix(responses, group = party_codes)
str(responses)
```

Note the messages printed by `as.response_matrix()`;
the function automatically handles unanimous items for us
(and lets us know which problematic items had to be discarded),
and specifies a prior mean of 0 for any groups not specified.
The default `prior_means` argument specifies that groups labelled
`100` (Voteview's code for Democrats) have a prior mean of -1,
those labelled `200` (Voteview's code for Republicans) have a prior mean of 1,
and those labelled `0` have a prior mean of 0.
We encountered an unspecified group, `328` (Voteview's code for Independents),
and by default specified a prior mean of 0 for that group.

Now we can run our sampler:

```{r run_sampler, eval=FALSE}
set.seed(1119) # For reproducibility
samples <- gpirtMCMC(responses, sample_iterations = 5000, burn_iterations = 0)
str(samples)
```

Then we can get estimates of our ideology parameters without ever having to
assume the precise functional form between Senators' ideology and their roll
call vote responses:

```{r plot_estimates, eval=FALSE}
theta_hat <- apply(samples[["theta"]], 2, mean)
nominate1 <- member_data$nominate_dim1[match(rownames(responses), icpsrs)]
plot_colors <- ifelse(party_codes == 100, "#0072b280",
                      ifelse(party_codes == 200, "#d55e0080", "#80808080"))
plot(nominate1, theta_hat, pch = 19, col = plot_colors,
     ylab = "", xlab = "", xaxt = "n", yaxt = "n")
axis(1, tick = FALSE, line = -0.75)
axis(2, tick = FALSE, line = -0.75)
mtext("NOMINATE-2D Dimension 1 (Voteview)", 1, 1.5)
mtext(expression(hat(theta)), 2, 1.5)
text(min(nominate1), max(theta_hat), adj = 0:1,
     paste("Correlation:", round(cor(theta_hat, nominate1), 3)))
legend("bottomright", bty = "n", pch = 19, col = unique(plot_colors),
       legend = c("Republican", "Democrat", "Independent"))
```

# References
